Reproducible Research: Peer Assessment 1
========================================
ASSUMPTIONS: The working directory contains the data file "activity.csv

## Load required libraries
```{r echo = TRUE}
require(sqldf)
require(tcltk)
require(lattice)
```
## Load raw data
dfRaw: dim=17568,3
       Steps:[int] Steps in Interval
     , Date:[Date] "2012-10-01".."2012-11-30"
     , Interval:[Integer] 5 minute interval 1205 means 12:05  
```{r}
unzip("activity.zip")
dfRaw <- read.csv("activity.csv", colClasses=c("integer", "Date", "integer"), col.names=c("Steps", "Date", "Interval"))
```
## Data transformations
dfStepsByDate: dim=53,2
      Steps:[int] Average Steps per day
    , Date:[Date] dim=53,2. "2012-10-02".."2012-11-29""
Filter out rows with missing (NA) "Steps".
```{r}
dfStepsByDate <- sqldf("select sum(Steps) as Steps, Date from dfRaw where Steps is not null group by Date;")
```
### What is mean total number of steps taken per day?
1.    Make a histogram of the total number of steps taken each day
```{r fig.width=7, fig.height=6}
p <- tapply(dfStepsByDate$Steps, dfStepsByDate$Date, FUN=sum)
# array (int, chr) # dim=53, 2012-10-02 .. 2012-11-29

barplot(p, main="Steps per Day", xlab="Date", ylab="Steps")
```

2.    Calculate and report the mean and median total number of steps taken per day
```{r}
dfSummary <- sqldf("SELECT AVG(Steps) AS MeanStepsPerDay, MEDIAN(Steps) AS MedianStepsPerDay FROM dfStepsByDate;")
summaryFormat <- "Statistics of Steps per day when ignoring rows with missing values:"
summaryFormat <- paste(summaryFormat, "Mean=%9.2f", sep="\n")
summaryFormat <- paste(summaryFormat, "Median=%d", sep="\n")
cat(sprintf(summaryFormat, dfSummary$MeanStepsPerDay, dfSummary$MedianStepsPerDay))
# Mean=10766.19, Median=10765 about 37 per interval
```
### What is the average daily activity pattern?
1.    Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
dfStepsByInterval <- sqldf("select avg(Steps) as Steps, Interval from dfRaw where Steps is not null group by Interval;") # dim=288,2
with(dfStepsByInterval, plot(Interval, Steps, main="Steps per 5 minute Interval", type="l", xlab="Time of Day", ylab="Average Steps per Day"))
```
2.    Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
x <- dfStepsByInterval[dfStepsByInterval$Steps == max(dfStepsByInterval$Steps),]
# Max Steps=206 at 08:35
summaryFormat <- "The 5-minute interval=%02d:%02d has the most steps=%d"
cat(sprintf(summaryFormat, x$Interval %/% 100, x$Interval %% 100, x$Steps))
```
## Imputing missing values
1.    Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
sum(is.na(dfRaw$Steps)) # Records with a missing Steps. 2304 out of 17568
sum(!complete.cases(dfRaw)) # A more accurate method that checks all columns.
```
2.    Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or <b>the mean for that 5-minute interval </b>, etc.

Fill missing 'Steps' with the mean over all days for the corresponding 5 minute interval.

3.    Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
dfTidy <- dfRaw

logArrMissing <- is.na(dfTidy$Steps) # Logical array 1:17568
intArrMissingRowIndex <- which(logArrMissing) # Integer array 1:2304
for (rowIndex in intArrMissingRowIndex) {
    Interval <- dfTidy[rowIndex, "Interval"]
    meanSteps <- dfStepsByInterval$Steps[dfStepsByInterval$Interval == Interval]
    dfTidy[rowIndex, "Steps"] <- meanSteps
}
```
4.    Make a histogram of the total number of steps taken each day and Calculate 
and report the <b>mean</b> and <b>median</b> total number of steps taken per day. 
```{r}
dfTidyStepsByInterval <- sqldf("SELECT avg(Steps) AS Steps, Interval FROM dfTidy GROUP BY Interval;")

with(dfTidyStepsByInterval, plot(Interval, Steps, main="Steps per 5 minute Interval. Missing=Interval mean.", type="l", xlab="Time of Day", ylab="Average Steps per Day"))
```
Do these values differ from the estimates from the first part of the assignment?
```{r}
sum(abs(dfTidyStepsByInterval$Steps - dfStepsByInterval$Steps)) # 0
```
What is the impact of imputing missing data on the estimates of the total daily number of steps?

There is NO difference when approximating missing values for 'total daily number of steps'
Comparing dfTidyStepsByInterval$Steps, dfStepsByInterval$Steps shows no difference. This was not expected!

## Are there differences in activity patterns between weekdays and weekends?
1.    Create a new factor variable in the dataset with two levels ??? "weekday" and "weekend"" 
indicating whether a given date is a weekday or weekend day.
```{r}
dfTidy$DayType <- as.factor(ifelse((weekdays(as.Date(dfTidy$Date)) %in% c('Saturday','Sunday')), "Weekend", "Weekday"))
```
2.    Make a panel plot containing a time series plot (i.e. type = "l") 
of the 5-minute interval (x-axis) and the average number of steps taken, 
averaged across all weekday days or weekend days (y-axis). 
```{r}
dfStepsByIntervalDayType <- sqldf("SELECT avg(Steps) AS Steps, DayType, Interval FROM dfTidy GROUP BY DayType, Interval;") # dim=288,2
xyplot(Steps ~ Interval|DayType, data=dfStepsByIntervalDayType, type='l', layout=c(1,2), ylab="Number of steps")
```
